---
title: "Flight Delay Prediction"
subtitle: "36-673 Final Project"
author: "Ellie Huang (elliehua) and Nikita Jayaprakash (njayapra)"
date: "2026-04-22"
output: pdf_document
---

# Data Prep

## Load Libraries and Data

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, cache = TRUE, warning = FALSE, message = FALSE)

library(tidyverse)
library(adabag)
library(caret)
library(e1071)
library(tidymodels)
library(here)
library(janitor)
library(lubridate)
library(pROC)
library(randomForest)
library(skimr)
library(scales)
library(forcats)
library(ranger)
library(glmnet)
library(xgboost)
```

```{r}
# Load data
flights2023 <- read_csv(here("data", "flights2023.csv")) |>
 clean_names()

flights2024 <- read_csv(here("data", "flights2024.csv")) |>
 clean_names()

test <- read_csv(here("data", "flights2025.csv")) |>
 clean_names()

test_visible <- read_csv(here("data", "flights2025_visible.csv")) |>
 clean_names() |>
  mutate(data_year = 2025)

test_guess <- read_csv(here("data", "flights2025_guess.csv")) |>
 clean_names() |>
  mutate(data_year = 2025)

# Build training set
train <- bind_rows(
 flights2023 |> mutate(data_year = 2023),
 flights2024 |> mutate(data_year = 2024)
)
```

## Clean and Filter Data

```{r}
# Apply the same cleaning steps to each data frame so train and test sets are consistent
clean_flights <- function(df) {
 df |>
 clean_names() |>
 mutate(
 fl_date = as.Date(fl_date),
 dep_del15 = as.numeric(dep_del15),
 op_unique_carrier = as.factor(op_unique_carrier),
 origin = as.factor(origin),
 dest = as.factor(dest),
 sched_dep_hour = floor(crs_dep_time / 100),
 sched_arr_hour = floor(crs_arr_time / 100),
 month = as.factor(month),
 day_of_week = as.factor(day_of_week)
 )
}

train <- clean_flights(train)
test <- clean_flights(test)
test_visible <- clean_flights(test_visible)
test_guess <- clean_flights(test_guess)

# Restrict to departures from PIT
train <- train |>
  filter(origin == "PIT")

test <- test |>
  filter(origin == "PIT")

test_visible <- test_visible |>
  filter(origin == "PIT")

test_guess <- test_guess |>
  filter(origin == "PIT")

# Stable row key for joining full 2025 schedule features back to visible/guess subsets
make_flight_key <- function(df) {
  df |>
    mutate(
      flight_key = paste(
        fl_date,
        op_unique_carrier,
        dest,
        crs_dep_time,
        flights,
        sep = "__"
      )
    )
}

train <- make_flight_key(train)
test <- make_flight_key(test)
test_visible <- make_flight_key(test_visible)
test_guess <- make_flight_key(test_guess)

# Quick check of target distribution
table(train$dep_del15, useNA = "ifany")
```

We focus on Pittsburgh departures because that is the airport represented in the prediction task. We also use `sched_dep_hour` rather than actual departure time, since actual departure time is not available for the flights we need to predict.

# Exploratory Data Analysis

The goal of the EDA is not just to describe the data, but to identify patterns that are both predictive and available before departure. In particular, we want to see whether delay risk depends in a stable way on scheduling, carrier, route, seasonality, and same-day operating conditions.

## Missingness

```{r}
# Missingness in the main timing and target fields
colSums(is.na(train[, c("dep_del15", "dep_time", "crs_dep_time")]))

# Variables with nonzero missingness
train |>
 summarise(across(everything(), ~mean(is.na(.)))) |>
 pivot_longer(everything(), names_to = "variable", values_to = "pct_missing") |>
 filter(pct_missing > 0) |>
 arrange(desc(pct_missing)) |>
 print(n = 30)
```

```{r}
# Missingness among key pre-departure predictors
key_predictors <- c(
 "dep_del15", "crs_dep_time", "sched_dep_hour", "op_unique_carrier",
 "dest", "distance", "month", "day_of_week", "flights"
)

key_predictors <- intersect(key_predictors, names(train))

train |>
 summarise(across(all_of(key_predictors), ~mean(is.na(.)))) |>
 pivot_longer(everything(), names_to = "variable", values_to = "pct_missing") |>
 mutate(pct_missing = percent(pct_missing, accuracy = 0.1))
```

```{r}
# Check whether missing targets line up with likely cancellations
train |>
 mutate(cancelled_proxy = is.na(dep_time) & is.na(dep_del15)) |>
 count(cancelled_proxy, missing_target = is.na(dep_del15))
```

Missing target values align exactly with likely cancellations, so this is not unexplained missingness. We therefore treat these rows as cancelled flights and exclude them from the delay prediction task.

## Departure Delay Distribution

```{r}
# Target distribution
train |>
 count(dep_del15)

train |>
 ggplot(aes(x = factor(dep_del15, labels = c("on time", "delayed")))) +
 geom_bar() +
 labs(
 title = "Departure delay distribution",
 x = NULL,
 y = "count"
 )
```

The classes are roughly 80/20 on-time versus delayed. This imbalance is worth keeping in mind, but it is not too concerning since our main evaluation metric is AUC rather than raw accuracy.

## Delay Rate by Carrier

```{r}
carrier_delay_tbl <- train |>
 filter(!is.na(dep_del15)) |>
 group_by(op_unique_carrier) |>
 summarise(
 delay_rate = mean(dep_del15),
 n = n(),
 .groups = "drop"
 ) |>
 arrange(desc(delay_rate))

carrier_delay_tbl |>
 ggplot(aes(x = reorder(op_unique_carrier, delay_rate), y = delay_rate)) +
 geom_col() +
 coord_flip() +
 scale_y_continuous(labels = percent_format()) +
 labs(
 title = "Delay rate by carrier",
 x = "carrier",
 y = "proportion delayed >= 15 min"
 ) +
 theme_minimal()
```

Delay rates vary meaningfully across carriers, so carrier should be a useful predictor. At the same time, some carriers have much lower flight counts at PIT than others, so raw group means will be noisy. That motivates using smoothed historical delay rates later in feature engineering instead of unsmoothed averages.

## Delay Rate by Scheduled Departure Hour

```{r}
# Distribution of scheduled departure times
train |>
 ggplot(aes(x = sched_dep_hour)) +
 geom_histogram(binwidth = 1, boundary = 0, closed = "left") +
 scale_x_continuous(breaks = 0:23) +
 labs(
 title = "Distribution of scheduled departure hours",
 x = "scheduled departure hour",
 y = "count"
 ) +
 theme_minimal()
```

```{r}
hour_delay_tbl <- train |>
 filter(!is.na(dep_del15), !is.na(sched_dep_hour)) |>
 group_by(sched_dep_hour) |>
 summarise(
 delay_rate = mean(dep_del15),
 n = n(),
 .groups = "drop"
 )

hour_delay_tbl |>
 ggplot(aes(x = sched_dep_hour, y = delay_rate)) +
 geom_line() +
 geom_point() +
 scale_x_continuous(breaks = 0:23) +
 scale_y_continuous(labels = percent_format()) +
 labs(
 title = "Delay rate by scheduled departure hour",
 x = "scheduled departure hour",
 y = "proportion delayed >= 15 min"
 ) +
 theme_minimal()
```

Delay rates rise steadily over the course of the day and are highest in the evening. This fits the idea of cascading delays: early flights start the day with fewer disruptions from previous legs, while later flights are more exposed to effects from earlier problems. This makes scheduled departure hour one of the strongest single predictors in the dataset.

## Carrier-by-Hour Interaction

```{r}
carrier_hour_tbl <- train |>
 filter(!is.na(dep_del15), !is.na(sched_dep_hour)) |>
 group_by(op_unique_carrier, sched_dep_hour) |>
 summarise(
 delay_rate = mean(dep_del15),
 n = n(),
 .groups = "drop"
 ) |>
 filter(n >= 20)

carrier_hour_tbl |>
 ggplot(
 aes(
 x = sched_dep_hour,
 y = fct_reorder(op_unique_carrier, delay_rate, .fun = mean),
 fill = delay_rate
 )
 ) +
 geom_tile(color = "white") +
 scale_fill_gradient(labels = percent_format(accuracy = 1)) +
 labs(
 title = "Delay rate by carrier and scheduled departure hour",
 subtitle = "Carrier differences are not constant across the day",
 x = "scheduled departure hour",
 y = "carrier",
 fill = "delay rate"
 ) +
 theme_minimal()
```

The carrier effect is not the same at every hour. Some carriers remain fairly stable across the day, while others become much more delay-prone in the evening. This suggests that a model with only main effects may miss meaningful structure, so either an explicit carrier-by-hour feature or a nonlinear model should help.

## Seasonality and Calendar Effects

```{r}
# Delay rate by month
train |>
 filter(!is.na(dep_del15)) |>
 group_by(month) |>
 summarise(
 delay_rate = mean(dep_del15),
 .groups = "drop"
 ) |>
 ggplot(aes(x = month, y = delay_rate, group = 1)) +
 geom_line() +
 geom_point() +
 scale_y_continuous(labels = percent_format()) +
 labs(
 title = "Delay rate by month",
 x = "month",
 y = "proportion delayed >= 15 min"
 ) +
 theme_minimal()
```

```{r}
# Delay rate by day of week
train |>
 filter(!is.na(dep_del15)) |>
 group_by(day_of_week) |>
 summarise(
 delay_rate = mean(dep_del15),
 .groups = "drop"
 ) |>
 ggplot(aes(x = day_of_week, y = delay_rate, group = 1)) +
 geom_line() +
 geom_point() +
 scale_y_continuous(labels = percent_format()) +
 labs(
 title = "Delay rate by day of week",
 x = NULL,
 y = "proportion delayed >= 15 min"
 ) +
 theme_minimal()
```

```{r}
# Delay rate by calendar date, aligned to a common year for display
train |>
 filter(!is.na(dep_del15)) |>
 mutate(plot_date = make_date(year = 2010, month = as.integer(as.character(month)), day = day_of_month)) |>
 group_by(plot_date) |>
 summarise(
 delay_rate = mean(dep_del15),
 .groups = "drop"
 ) |>
 ggplot(aes(x = plot_date, y = delay_rate)) +
 geom_line() +
 scale_x_date(date_labels = "%b", date_breaks = "1 month") +
 scale_y_continuous(labels = percent_format()) +
 labs(
 title = "Daily delay rate across the calendar year",
 x = NULL,
 y = "proportion delayed >= 15 min"
 ) +
 theme_minimal()
```

Delays are worse in the summer months and in December, which is consistent with heavier travel volume and more weather-related disruption. The day-of-week pattern is milder, but it still suggests that weekly operating patterns contain some signal. The daily calendar plot also shows that these seasonal peaks are not spread evenly across the month, which supports adding a simple calendar feature rather than relying only on month.

## Destination and Distance Effects

```{r}
top_dests <- train |>
 count(dest, sort = TRUE) |>
 slice_head(n = 15) |>
 pull(dest)

train |>
 filter(!is.na(dep_del15), dest %in% top_dests) |>
 group_by(dest) |>
 summarise(
 delay_rate = mean(dep_del15),
 n = n(),
 .groups = "drop"
 ) |>
 arrange(delay_rate) |>
 ggplot(aes(x = reorder(dest, delay_rate), y = delay_rate)) +
 geom_col() +
 coord_flip() +
 scale_y_continuous(labels = percent_format()) +
 labs(
 title = "Delay rate among the busiest PIT destinations",
 x = "destination",
 y = "proportion delayed >= 15 min"
 ) +
 theme_minimal()
```

```{r}
train <- train |>
 mutate(
 distance_bucket = case_when(
 distance > 600 ~ ">600",
 distance > 300 ~ "301-600",
 TRUE ~ "<=300"
 ),
 distance_bucket = factor(
 distance_bucket,
 levels = c("<=300", "301-600", ">600")
 )
 )

train |>
 filter(!is.na(dep_del15), !is.na(distance_bucket)) |>
 group_by(distance_bucket) |>
 summarise(
 delay_rate = mean(dep_del15),
 n = n(),
 .groups = "drop"
 ) |>
 ggplot(aes(x = distance_bucket, y = delay_rate)) +
 geom_col() +
 scale_y_continuous(labels = percent_format()) +
 labs(
 title = "Longer routes have higher delay rates",
 x = "distance bucket",
 y = "proportion delayed >= 15 min"
 ) +
 theme_minimal()
```

Delay rates differ across destinations, even among the busiest PIT routes, which suggests that route context adds information beyond carrier and time of day alone. Longer routes also show higher delay rates on average, so distance should remain in the model and may be easier to capture through both a raw continuous version and a bucketed factor version

## High-Delay Days

```{r}
bad_days_tbl <- train |>
 filter(!is.na(dep_del15)) |>
 group_by(fl_date) |>
 summarise(
 delay_rate = mean(dep_del15),
 n = n(),
 .groups = "drop"
 ) |>
 filter(n >= 10) |>
 arrange(desc(delay_rate))

bad_days_tbl |>
 slice_head(n = 20) |>
 ggplot(aes(x = reorder(as.character(fl_date), delay_rate), y = delay_rate)) +
 geom_col() +
 coord_flip() +
 scale_y_continuous(labels = percent_format()) +
 labs(
 title = "A small number of dates have very high delay rates",
 x = "date",
 y = "proportion delayed >= 15 min"
 ) +
 theme_minimal()
```

A small number of days have unusually high delay rates, which suggests that there are broader operating shocks affecting many flights at once. That motivates adding dynamic same-day features based on earlier visible flights, along with a historical calendar-day disruption feature learned from the training data.

# Feature Engineering

The EDA suggests that delay risk depends on a mix of stable historical tendencies and day-specific operating conditions. Based on those patterns, we engineer features that capture time-of-day effects, carrier and route history, route distance, seasonality, and same-day disruption.

## Basic Features

```{r}
add_basic_features <- function(df) {
  df |>
    mutate(
      sched_dep_minute = crs_dep_time %% 100,
      sched_dep_mins_since_midnight = sched_dep_hour * 60 + sched_dep_minute,

      hour_bucket = case_when(
        sched_dep_hour <= 8  ~ "early_morning",
        sched_dep_hour <= 12 ~ "morning",
        sched_dep_hour <= 17 ~ "afternoon",
        sched_dep_hour <= 21 ~ "evening",
        TRUE ~ "late_night"
      ),
      hour_bucket = factor(
        hour_bucket,
        levels = c("early_morning", "morning", "afternoon", "evening", "late_night")
      ),

      peak_season = if_else(as.integer(as.character(month)) %in% c(6, 7, 8, 12), 1L, 0L),
      is_evening = if_else(sched_dep_hour >= 18, 1L, 0L),
      is_morning = if_else(sched_dep_hour <= 8, 1L, 0L),
      is_weekend = if_else(as.integer(as.character(day_of_week)) %in% c(6, 7), 1L, 0L),

      distance_bucket = case_when(
        distance > 600 ~ ">600",
        distance > 300 ~ "301-600",
        TRUE ~ "<=300"
      ),
      distance_bucket = factor(
        distance_bucket,
        levels = c("<=300", "301-600", ">600")
      ),

      long_flight = if_else(distance > 600, 1L, 0L),
      evening_long_flight = if_else(distance > 600 & sched_dep_hour >= 18, 1L, 0L),

      carrier_hour = paste(op_unique_carrier, sched_dep_hour, sep = "_"),
      route = paste(op_unique_carrier, dest, sep = "_")
    )
}
```

These features directly reflect the strongest patterns from the EDA. In particular, `peak_season`, the morning and evening indicators, the distance bucket, and the carrier-by-hour combination all transform observed structure in the data into modeling variables.

```{r}
add_schedule_density_features <- function(df) {
  df |>
    arrange(fl_date, crs_dep_time) |>
    group_by(fl_date) |>
    mutate(
      day_total_departures = n(),
      scheduled_position_today = row_number() - 1,
      day_progress = if_else(day_total_departures > 1,
                             scheduled_position_today / (day_total_departures - 1),
                             0)
    ) |>
    group_by(fl_date, sched_dep_hour) |>
    mutate(
      hour_total_departures = n(),
      scheduled_position_hour = row_number() - 1
    ) |>
    ungroup() |>
    mutate(
      hour_share_of_day = hour_total_departures / pmax(day_total_departures, 1)
    )
}

# Train features
train <- train |>
  add_basic_features() |>
  add_schedule_density_features()

# Compute schedule-based features on the full 2025 PIT departure file
test_full_fe <- test |>
  add_basic_features() |>
  add_schedule_density_features()

# Then split back into visible / guess using the stable key
test_visible <- test_full_fe |>
  semi_join(test_visible |> select(flight_key), by = "flight_key")

test_guess <- test_full_fe |>
  semi_join(test_guess |> select(flight_key), by = "flight_key")
```

## Smoothed Historical Delay Rates

Raw group means are noisy, especially for carriers and routes with few flights at PIT. We use James-Stein style shrinkage: each group's rate is a weighted blend of its observed mean and the global delay rate, with the weight determined by how many observations the group has. Groups with lots of data are trusted; groups with little data are pulled toward the global mean.

```{r}
global_delay_rate <- mean(train$dep_del15, na.rm = TRUE)

make_smoothed_rates <- function(df, group_var, target = "dep_del15", m = 100) {
  group_sym <- rlang::sym(group_var)
  target_sym <- rlang::sym(target)

  df |>
    filter(!is.na(!!target_sym)) |>
    group_by(!!group_sym) |>
    summarise(
      n = n(),
      mean_target = mean(!!target_sym),
      .groups = "drop"
    ) |>
    mutate(
      smoothed_rate = (n * mean_target + m * global_delay_rate) / (n + m)
    ) |>
    select(!!group_sym, smoothed_rate, n)
}
```

```{r}
# Route = carrier + dest; more specific than either alone
# Stable carrier and destination rates
carrier_rates <- make_smoothed_rates(train, "op_unique_carrier", m = 200) |>
  rename(
    carrier_delay_rate_smooth = smoothed_rate,
    carrier_n = n
  )

dest_rates <- make_smoothed_rates(train, "dest", m = 200) |>
  rename(
    dest_delay_rate_smooth = smoothed_rate,
    dest_n = n
  )
```

```{r}
# Out-of-fold (OOF) encoding for fine-grained features
# carrier_hour, route, and flight_num are granular enough that computing them on the full
# training set leaks label info into val rows. OOF fixes this: each row's rate is computed
# from the other 4 folds only. Test sets use full-train lookups since their labels are never seen.
set.seed(100)
n_folds <- 5
fold_id <- sample(rep(1:n_folds, length.out = nrow(train)))

make_oof_rates <- function(df, group_var, target = "dep_del15", m = 100) {
  group_sym <- rlang::sym(group_var)
  target_sym <- rlang::sym(target)
  global <- mean(df[[target]], na.rm = TRUE)

  result <- numeric(nrow(df))
  for (k in 1:n_folds) {
    in_fold <- which(fold_id == k)
    out_fold <- which(fold_id != k)

    lookup <- df[out_fold, ] |>
      filter(!is.na(!!target_sym)) |>
      group_by(!!group_sym) |>
      summarise(
        n = n(),
        mean_t = mean(!!target_sym),
        .groups = "drop"
      ) |>
      mutate(rate = (n * mean_t + m * global) / (n + m))

    merged <- df[in_fold, ] |>
      select(!!group_sym) |>
      left_join(lookup |> select(!!group_sym, rate), by = group_var)

    result[in_fold] <- coalesce(merged$rate, global)
  }

  result
}

train <- train |>
  mutate(
    week_of_year = as.character(week(fl_date))
  )

test_visible <- test_visible |>
  mutate(
    week_of_year = as.character(week(fl_date))
  )

test_guess <- test_guess |>
  mutate(
    week_of_year = as.character(week(fl_date))
  )

train$carrier_hour_delay_rate_smooth <- make_oof_rates(train, "carrier_hour", m = 100)
train$route_delay_rate_smooth        <- make_oof_rates(train, "route", m = 150)
train$flight_num_delay_rate_smooth   <- make_oof_rates(train, "flights", m = 150)
train$week_delay_rate_smooth         <- make_oof_rates(train, "week_of_year", m = 50)

# Full-train lookups for test sets
carrier_hour_rates <- make_smoothed_rates(train, "carrier_hour", m = 100) |>
  rename(
    carrier_hour_delay_rate_smooth = smoothed_rate,
    carrier_hour_n = n
  )

route_rates <- make_smoothed_rates(train, "route", m = 150) |>
  rename(
    route_delay_rate_smooth = smoothed_rate,
    route_n = n
  )

flight_num_rates <- make_smoothed_rates(train, "flights", m = 150) |>
  rename(
    flight_num_delay_rate_smooth = smoothed_rate,
    flight_num_n = n
  )

week_rates <- make_smoothed_rates(train, "week_of_year", m = 50) |>
  rename(
    week_delay_rate_smooth = smoothed_rate,
    week_n = n
  )
```

## Apply Historical Features to Train and Test

```{r}
# This joins both the rates and the support counts, and creates log-count features that can help trees and linear models
apply_rate_features <- function(df, is_train = FALSE) {
  df <- df |>
    left_join(carrier_rates, by = "op_unique_carrier") |>
    left_join(dest_rates, by = "dest") |>
    mutate(
      carrier_delay_rate_smooth = coalesce(carrier_delay_rate_smooth, global_delay_rate),
      dest_delay_rate_smooth = coalesce(dest_delay_rate_smooth, global_delay_rate)
    )

  if (!is_train) {
    df <- df |>
      left_join(carrier_hour_rates, by = "carrier_hour") |>
      left_join(route_rates, by = "route") |>
      left_join(flight_num_rates, by = "flights") |>
      left_join(week_rates, by = "week_of_year") |>
      mutate(
        carrier_hour_delay_rate_smooth = coalesce(carrier_hour_delay_rate_smooth, global_delay_rate),
        route_delay_rate_smooth = coalesce(route_delay_rate_smooth, global_delay_rate),
        flight_num_delay_rate_smooth = coalesce(flight_num_delay_rate_smooth, global_delay_rate),
        week_delay_rate_smooth = coalesce(week_delay_rate_smooth, global_delay_rate)
      )
  } else {
    # Add support counts for train rows too
    df <- df |>
      left_join(carrier_hour_rates |> select(carrier_hour, carrier_hour_n), by = "carrier_hour") |>
      left_join(route_rates |> select(route, route_n), by = "route") |>
      left_join(flight_num_rates |> select(flights, flight_num_n), by = "flights") |>
      left_join(week_rates |> select(week_of_year, week_n), by = "week_of_year")
  }

  df |>
    mutate(
      carrier_n = coalesce(carrier_n, 0),
      dest_n = coalesce(dest_n, 0),
      carrier_hour_n = coalesce(carrier_hour_n, 0),
      route_n = coalesce(route_n, 0),
      flight_num_n = coalesce(flight_num_n, 0),
      week_n = coalesce(week_n, 0),

      log_carrier_n = log1p(carrier_n),
      log_dest_n = log1p(dest_n),
      log_carrier_hour_n = log1p(carrier_hour_n),
      log_route_n = log1p(route_n),
      log_flight_num_n = log1p(flight_num_n)
    )
}

train_fe <- apply_rate_features(train, is_train = TRUE)
test_visible_fe <- apply_rate_features(test_visible, is_train = FALSE)
test_guess_fe <- apply_rate_features(test_guess, is_train = FALSE)
```

## Same-Day Disruption Features

```{r}
# Names now match across train, visible, and guess, and it adds a stronger “pressure” feature.
add_running_day_features <- function(df) {
  df |>
    arrange(fl_date, crs_dep_time) |>
    group_by(fl_date) |>
    mutate(
      before_today = row_number() - 1,
      delays_before_today = lag(cumsum(replace_na(dep_del15, 0)), default = 0),
      prior_same_day_delay_rate = if_else(
        before_today > 0,
        delays_before_today / before_today,
        0
      ),
      visible_count = before_today,
      visible_delay_rate = prior_same_day_delay_rate
    ) |>
    ungroup() |>
    mutate(
      disruption_pressure = visible_delay_rate * log1p(visible_count),
      schedule_pressure = log1p(day_total_departures) * day_progress
    )
}

train_fe <- add_running_day_features(train_fe)
test_visible_fe <- add_running_day_features(test_visible_fe)
```

```{r}
# For guess rows, use visible flights earlier that day
test_visible_day_summary <- test_visible_fe |>
  filter(!is.na(dep_del15)) |>
  group_by(fl_date) |>
  summarise(
    visible_count = n(),
    visible_delay_rate = mean(dep_del15),
    .groups = "drop"
  )

test_guess_fe <- test_guess_fe |>
  left_join(test_visible_day_summary, by = "fl_date") |>
  mutate(
    visible_count = coalesce(visible_count, 0),
    visible_delay_rate = coalesce(visible_delay_rate, global_delay_rate),
    disruption_pressure = visible_delay_rate * log1p(visible_count),
    schedule_pressure = log1p(day_total_departures) * day_progress
  )
```

For the guess set, we cannot directly use the unknown future labels. Instead, we summarize the flights that are already visible on that date and merge those day-level summaries into the flights we need to predict.

## Carrier and Destination Encodings

Raw `op_unique_carrier` and `dest` factors can't go directly into the model because 2025 may contain carriers or destinations not seen in training, which would produce NA columns in the model matrix. Instead we use three stable encodings: a carrier size bucket (large/medium/small by flight count), a geographic region grouping for destinations, and a hub airport flag.

```{r}
# Keeps your broader groupings, which are still useful
carrier_size <- train |>
  count(op_unique_carrier) |>
  mutate(
    carrier_size = case_when(
      n >= 5000 ~ "large",
      n >= 1000 ~ "medium",
      TRUE ~ "small"
    )
  ) |>
  select(op_unique_carrier, carrier_size)

hub_airports <- c(
  "ATL", "ORD", "DFW", "DEN", "CLT", "JFK", "LAX",
  "LGA", "EWR", "SFO", "SEA", "MIA", "BOS", "MSP", "DTW", "PHL"
)

dest_region <- train |>
  distinct(dest) |>
  mutate(
    dest_region = case_when(
      dest %in% c("BOS", "JFK", "LGA", "EWR", "PHL", "DCA", "IAD", "BWI") ~ "northeast",
      dest %in% c("ATL", "CLT", "MIA", "TPA", "MCO", "RDU", "BNA", "MEM") ~ "south",
      dest %in% c("ORD", "MDW", "DTW", "MSP", "CLE", "CMH", "IND", "MKE") ~ "midwest",
      dest %in% c("DFW", "DAL", "IAH", "HOU", "MSY", "SAT", "AUS") ~ "southwest",
      dest %in% c("DEN", "SLC", "PHX", "LAS", "ABQ") ~ "mountain",
      dest %in% c("LAX", "SFO", "SEA", "PDX", "SAN", "SJC", "OAK") ~ "west",
      TRUE ~ "other"
    )
  )

add_carrier_dest_features <- function(df) {
  df |>
    left_join(carrier_size, by = "op_unique_carrier") |>
    left_join(dest_region, by = "dest") |>
    mutate(
      carrier_size = coalesce(carrier_size, "small"),
      dest_region = coalesce(dest_region, "other"),
      is_hub_dest = if_else(dest %in% hub_airports, 1L, 0L),
      carrier_size = factor(carrier_size, levels = c("small", "medium", "large")),
      dest_region = factor(dest_region),
      high_disruption_calendar_day = if_else(
        week_delay_rate_smooth >= quantile(week_delay_rate_smooth, 0.90, na.rm = TRUE),
        1L,
        0L
      )
    )
}

train_fe <- add_carrier_dest_features(train_fe)
test_visible_fe <- add_carrier_dest_features(test_visible_fe)
test_guess_fe <- add_carrier_dest_features(test_guess_fe)
```

## Final Modeling Data Frames

```{r}
model_vars <- c(
  "dep_del15",

  # Time / schedule
  "sched_dep_hour",
  "sched_dep_mins_since_midnight",
  "hour_bucket",
  "month",
  "day_of_week",
  "peak_season",
  "is_evening",
  "is_morning",
  "is_weekend",

  # Distance / route shape
  "distance",
  "distance_bucket",
  "long_flight",
  "evening_long_flight",

  # Schedule congestion
  "day_total_departures",
  "day_progress",
  "hour_total_departures",
  "schedule_pressure",

  # Carrier / destination encodings
  "carrier_size",
  "dest_region",
  "is_hub_dest",

  # Historical delay tendency
  "carrier_delay_rate_smooth",
  "carrier_hour_delay_rate_smooth",
  "route_delay_rate_smooth",

  # Support / reliability
  "log_carrier_n",
  "log_carrier_hour_n",
  "log_route_n",

  # Same-day disruption
  "prior_same_day_delay_rate",
  "before_today",
  "disruption_pressure",

  # Calendar disruption
  "high_disruption_calendar_day"
)

train_model_df <- train_fe |>
  select(any_of(model_vars)) |>
  filter(!is.na(dep_del15))

test_visible_model_df <- test_visible_fe |>
  select(any_of(model_vars))

test_guess_model_df <- test_guess_fe |>
  select(any_of(model_vars))
```

At this point, `train_model_df` contains the response and engineered predictors for model fitting, while `test_guess_model_df` contains the same predictors for the flights we need to score. The final feature set reflects the main conclusions from the EDA: time of day is a major driver, carrier, destination, route are important, and same-day disruption adds useful information.

# Modelling

## Setup

```{r}
# Helper to align model matrices between train and test so factor levels always match.
# Adds zero columns for any level in train not seen in test, drops extras.
make_aligned_matrix <- function(formula, train_df, test_df) {
  x_train <- model.matrix(formula, data = train_df)[, -1, drop = FALSE]
  x_test  <- model.matrix(formula, data = test_df)[, -1, drop = FALSE]

  missing_cols <- setdiff(colnames(x_train), colnames(x_test))
  if (length(missing_cols) > 0) {
    for (col in missing_cols) {
      x_test <- cbind(x_test, 0)
      colnames(x_test)[ncol(x_test)] <- col
    }
  }

  extra_cols <- setdiff(colnames(x_test), colnames(x_train))
  if (length(extra_cols) > 0) {
    x_test <- x_test[, !(colnames(x_test) %in% extra_cols), drop = FALSE]
  }

  x_test <- x_test[, colnames(x_train), drop = FALSE]

  list(
    x_train = x_train,
    x_test = x_test
  )
}
```

```{r}
# Helper to align factor levels between a new data frame and the training split.
# Needed for randomForest which requires the same factor levels at predict time.
align_to_training <- function(new_df, train_df) {
  common_cols <- intersect(names(train_df), names(new_df))

  for (col in common_cols) {
    # align factor columns
    if (is.factor(train_df[[col]])) {
      new_df[[col]] <- factor(as.character(new_df[[col]]),
                              levels = levels(train_df[[col]]))
    }

    # align numeric/integer columns
    if (is.numeric(train_df[[col]]) && !is.numeric(new_df[[col]])) {
      new_df[[col]] <- as.numeric(new_df[[col]])
    }

    if (is.integer(train_df[[col]]) && !is.integer(new_df[[col]])) {
      new_df[[col]] <- as.integer(new_df[[col]])
    }
  }

  new_df
}
```

```{r}
# 80/20 train/val split for internal model evaluation
set.seed(100)

n <- nrow(train_model_df)
train_idx <- sample(1:n, size = 0.8 * n)

train_split <- train_model_df[train_idx, ]
test_split  <- train_model_df[-train_idx, ]
```

```{r}
# AUC and accuracy tracker — appends one row per model
compute_auc <- function(auc_df = NULL, train_y, train_prob,
                        test_y, test_prob, model_name) {
  auc_train <- roc(train_y, train_prob) |> auc()
  auc_test <- roc(test_y,  test_prob)  |> auc()

  accur_train_0.5 <- mean(train_y == (train_prob > 0.5))
  accur_test_0.5 <- mean(test_y  == (test_prob  > 0.5))

  accur_train_0.6 <- mean(train_y == (train_prob > 0.6))
  accur_test_0.6 <- mean(test_y  == (test_prob  > 0.6))

  df <- data.frame(
    model_name,
    auc_train,
    auc_test,
    accur_train_0.5,
    accur_test_0.5,
    accur_train_0.6,
    accur_test_0.6
  )

  if (is.null(auc_df)) {
    return(df)
  } else {
    return(rbind(auc_df, df))
  }
}
```

```{r}
y_train <- train_split$dep_del15
y_test  <- test_split$dep_del15

# Linear model matrix: poly(sched_dep_hour, 2) lets logistic/LASSO capture the curved hour effect
lm_mats <- make_aligned_matrix(
  dep_del15 ~ . + poly(sched_dep_hour, 2),
  train_split,
  test_split
)

x_train_lm <- lm_mats$x_train
x_test_lm <- lm_mats$x_test

# XGBoost matrix: raw features only — XGBoost finds nonlinearity through splits, not poly terms
xgb_mats <- make_aligned_matrix(
  dep_del15 ~ .,
  train_split,
  test_split
)

x_train_xgb <- xgb_mats$x_train
x_test_xgb <- xgb_mats$x_test

# Full-train matrix used for XGBoost CV tuning (avoids overfitting to the 80/20 split)
x_full_xgb <- model.matrix(dep_del15 ~ ., data = train_model_df)[, -1, drop = FALSE]
y_full <- train_model_df$dep_del15
```

## Logistic Regression

```{r}
log_reg <- glm(
  dep_del15 ~ . + poly(sched_dep_hour, 2),
  family = "binomial",
  data = train_split
)

train_prob <- predict(log_reg, type = "response", newdata = train_split)
test_prob  <- predict(log_reg, type = "response", newdata = test_split)

auc_df <- compute_auc(
  train_y = train_split$dep_del15,
  train_prob = train_prob,
  test_y = test_split$dep_del15,
  test_prob = test_prob,
  model_name = "Log Reg"
)
```

## LASSO

```{r}
cv_lasso <- cv.glmnet(
  x_train_lm,
  y_train,
  family = "binomial",
  alpha = 1
)

train_prob <- predict(
  cv_lasso,
  newx = x_train_lm,
  s = "lambda.min",
  type = "response"
) |>
  as.vector()

test_prob <- predict(
  cv_lasso,
  newx = x_test_lm,
  s = "lambda.min",
  type = "response"
) |>
  as.vector()

auc_df <- compute_auc(
  auc_df = auc_df,
  train_y = y_train,
  train_prob = train_prob,
  test_y = y_test,
  test_prob = test_prob,
  model_name = "Log Reg (LASSO)"
)
```

## XGBoost (tuned)

XGBoost with default params and 100 rounds leaves a lot of AUC on the table. We tune
the key parameters via cross-validation: `max_depth` controls tree complexity, `eta`
controls shrinkage, and `subsample`/`colsample_bytree` add randomization to reduce
overfitting. `nrounds` is selected by early stopping on the CV AUC curve.

```{r}
dtrain <- xgb.DMatrix(data = x_train_xgb, label = y_train)
dtest  <- xgb.DMatrix(data = x_test_xgb, label = y_test)
dfull  <- xgb.DMatrix(data = x_full_xgb, label = y_full)

xgb_grid <- expand.grid(
  max_depth        = c(3, 4, 5),
  eta              = c(0.05, 0.1),
  subsample        = c(0.7, 0.9),
  colsample_bytree = c(0.7, 0.9),
  min_child_weight = c(10, 30),
  gamma            = c(0, 1)
)

best_auc     <- 0
best_params  <- NULL
best_nrounds <- 100

for (i in seq_len(nrow(xgb_grid))) {
  params_i <- list(
    objective        = "binary:logistic",
    eval_metric      = "auc",
    max_depth        = xgb_grid$max_depth[i],
    eta              = xgb_grid$eta[i],
    subsample        = xgb_grid$subsample[i],
    colsample_bytree = xgb_grid$colsample_bytree[i],
    min_child_weight = xgb_grid$min_child_weight[i],
    gamma            = xgb_grid$gamma[i]
  )

  cv_result <- xgb.cv(
    params = params_i,
    data = dfull,
    nrounds = 500,
    nfold = 5,
    early_stopping_rounds = 30,
    verbose = 0
  )

  nrounds_i <- cv_result$best_iteration
  if (is.null(nrounds_i) || length(nrounds_i) == 0) {
    nrounds_i <- which.max(cv_result$evaluation_log$test_auc_mean)
  }

  best_i <- cv_result$evaluation_log$test_auc_mean[nrounds_i]

  if (best_i > best_auc) {
    best_auc     <- best_i
    best_params  <- params_i
    best_nrounds <- nrounds_i
  }
}

cat("Best CV AUC:", round(best_auc, 4), "| nrounds:", best_nrounds, "\n")

xgb_model <- xgb.train(
  params = best_params,
  data = dfull,
  nrounds = best_nrounds,
  verbose = 0
)

train_prob <- predict(xgb_model, dtrain)
test_prob  <- predict(xgb_model, dtest)

auc_df <- compute_auc(
  auc_df = auc_df,
  train_y = y_train,
  train_prob = train_prob,
  test_y = y_test,
  test_prob = test_prob,
  model_name = "XGBoost (tuned)"
)
```

## Random Forest

```{r}
train_split_rf <- train_split |>
  mutate(dep_del15 = as.factor(dep_del15))

test_split_rf <- test_split |>
  mutate(dep_del15 = as.factor(dep_del15))

rf_model <- randomForest(
  dep_del15 ~ .,
  data = train_split_rf,
  ntree = 300,
  importance = TRUE
)

train_prob <- predict(rf_model, newdata = train_split_rf, type = "prob")[, "1"]

test_split_rf_aligned <- align_to_training(test_split_rf, train_split_rf)

test_prob <- predict(rf_model, newdata = test_split_rf_aligned, type = "prob")[, "1"]

auc_df <- compute_auc(
  auc_df = auc_df,
  train_y = y_train,
  train_prob = train_prob,
  test_y = y_test,
  test_prob = test_prob,
  model_name = "Random Forest"
)
```

## Balanced Random Forest

```{r}
# Upweight the minority (delayed) class to counteract the 80/20 imbalance
class_counts <- table(train_split_rf$dep_del15)

class_weights <- as.numeric(
  sum(class_counts) / (length(class_counts) * class_counts)
)
names(class_weights) <- names(class_counts)

brf_model <- randomForest(
  dep_del15 ~ .,
  data = train_split_rf,
  ntree = 300,
  classwt = class_weights
)

train_prob <- predict(brf_model, newdata = train_split_rf, type = "prob")[, "1"]

test_split_rf_aligned <- align_to_training(test_split_rf, train_split_rf)

test_prob <- predict(brf_model, newdata = test_split_rf_aligned, type = "prob")[, "1"]

auc_df <- compute_auc(
  auc_df = auc_df,
  train_y = y_train,
  train_prob = train_prob,
  test_y = y_test,
  test_prob = test_prob,
  model_name = "Balanced RF"
)
```

## Ensembles

Simple averaging across all models dilutes the best performers. We compare: (1) the naive
average, (2) a weighted average that upweights XGBoost, which typically dominates on
tabular data, and (3) majority vote.

```{r}
preds_df <- data.frame(
  y_true = as.numeric(as.character(test_split$dep_del15)),
  log_reg = predict(log_reg, type = "response", newdata = test_split),
  lasso = predict(cv_lasso, newx = x_test_lm, s = "lambda.min", type = "response") |> as.vector(),
  xgboost = predict(xgb_model, dtest),
  rf = predict(rf_model,  newdata = test_split_rf, type = "prob")[, "1"],
  brf = predict(brf_model, newdata = test_split_rf, type = "prob")[, "1"]
)

# Simple average
preds_df$avg_prob <- rowMeans(preds_df[, c("log_reg", "lasso", "xgboost", "rf", "brf")])

auc_df <- compute_auc(
  auc_df = auc_df,
  train_y = preds_df$y_true,
  train_prob = preds_df$avg_prob,
  test_y = preds_df$y_true,
  test_prob = preds_df$avg_prob,
  model_name = "Ensemble (Avg)"
)

# Weighted average: XGBoost gets more weight since it's the strongest model on this data
preds_df$weighted_prob <- with(
  preds_df,
  0.50 * xgboost +
    0.15 * rf +
    0.10 * brf +
    0.15 * lasso +
    0.10 * log_reg
)

auc_df <- compute_auc(
  auc_df = auc_df,
  train_y = preds_df$y_true,
  train_prob = preds_df$weighted_prob,
  test_y = preds_df$y_true,
  test_prob = preds_df$weighted_prob,
  model_name = "Ensemble (Weighted)"
)

# Majority vote
pred_class <- preds_df[, c("log_reg", "lasso", "xgboost", "rf", "brf")] > 0.5
preds_df$majority_prob <- rowMeans(pred_class)

auc_df <- compute_auc(
  auc_df = auc_df,
  train_y = preds_df$y_true,
  train_prob = preds_df$majority_prob,
  test_y = preds_df$y_true,
  test_prob = preds_df$majority_prob,
  model_name = "Ensemble (Majority Vote)"
)
```

# Results - 2025 Visible

```{r}
# AUC and accuracy on the 2025 visible set (the portion of test data where we know the true labels)
compute_vis_auc <- function(auc_df = NULL, y, prob, model_name) {
  auc <- roc(y, prob) |> auc()
  accuracy <- mean(y == (prob > 0.5), na.rm = TRUE)

  df <- data.frame(model_name, auc, accuracy)

  if (is.null(auc_df)) {
    return(df)
  } else {
    return(rbind(auc_df, df))
  }
}
```

```{r}
test_visible_model_df_eval <- test_visible_model_df |>
  filter(!is.na(dep_del15))

y_true <- as.numeric(as.character(test_visible_model_df_eval$dep_del15))
y_vis <- test_visible_model_df_eval$dep_del15

# Build aligned matrices using the original training split as the reference
vis_lm_mats <- make_aligned_matrix(
  dep_del15 ~ . + poly(sched_dep_hour, 2),
  train_split,
  test_visible_model_df_eval
)

x_vis_lm <- vis_lm_mats$x_test

vis_xgb_mats <- make_aligned_matrix(
  dep_del15 ~ .,
  train_split,
  test_visible_model_df_eval
)

x_vis_xgb <- vis_xgb_mats$x_test
dvis <- xgb.DMatrix(data = x_vis_xgb, label = y_vis)

# Logistic / lasso / xgboost probabilities
log_reg_prob <- predict(log_reg, type = "response", newdata = test_visible_model_df_eval)

lasso_prob <- predict(
  cv_lasso,
  newx = x_vis_lm,
  s = "lambda.min",
  type = "response"
) |>
  as.vector()

xgb_prob <- predict(xgb_model, dvis)

# RF-compatible visible data (factor levels must match training)
test_visible_rf <- test_visible_model_df_eval |>
  mutate(dep_del15 = as.factor(dep_del15))

test_visible_rf <- align_to_training(test_visible_rf, train_split_rf)

rf_prob <- predict(
  rf_model,
  newdata = test_visible_rf,
  type = "prob"
)[, "1"]

brf_prob <- predict(
  brf_model,
  newdata = test_visible_rf,
  type = "prob"
)[, "1"]

avg_prob <- rowMeans(data.frame(log_reg_prob, lasso_prob, xgb_prob, rf_prob, brf_prob))

weighted_prob <- 0.50 * xgb_prob +
  0.15 * rf_prob +
  0.10 * brf_prob +
  0.15 * lasso_prob +
  0.10 * log_reg_prob

vis_auc <- compute_vis_auc(
  y = y_true,
  prob = log_reg_prob,
  model_name = "Log Reg"
)

vis_auc <- compute_vis_auc(
  vis_auc,
  y = y_true,
  prob = lasso_prob,
  model_name = "LASSO"
)

vis_auc <- compute_vis_auc(
  vis_auc,
  y = y_true,
  prob = xgb_prob,
  model_name = "XGBoost (tuned)"
)

vis_auc <- compute_vis_auc(
  vis_auc,
  y = y_true,
  prob = rf_prob,
  model_name = "RF"
)

vis_auc <- compute_vis_auc(
  vis_auc,
  y = y_true,
  prob = brf_prob,
  model_name = "Balanced RF"
)

vis_auc <- compute_vis_auc(
  vis_auc,
  y = y_true,
  prob = avg_prob,
  model_name = "Ensemble (Avg)"
)

vis_auc <- compute_vis_auc(
  vis_auc,
  y = y_true,
  prob = weighted_prob,
  model_name = "Ensemble (Weighted)"
)
```

```{r}
vis_auc |> arrange(desc(auc))
```

```{r}
vis_auc |> arrange(desc(accuracy))
```

## XGBoost Feature Importance

```{r}
xgb.importance(model = xgb_model) |>
  xgb.plot.importance(top_n = 15, main = "XGBoost feature importance (top 15)")
```

```{r}
# Prepare guess data for model.matrix — dep_del15 is all NA bc it's what we're predicting
# model.matrix can't handle factors with < 2 levels, so we clean first
prep_guess <- function(df, ref_df) {
  # drop the target
  df <- df |> select(-any_of("dep_del15"))
  ref_df <- ref_df |> select(-any_of("dep_del15"))

  # for each factor column, force its levels to match the reference (train_split)
  # so model.matrix sees the same level set and doesn't throw error on single-level factors
  for (col in names(df)) {
    if (is.factor(ref_df[[col]]) && col %in% names(df)) {
      df[[col]] <- factor(as.character(df[[col]]),
                          levels = levels(ref_df[[col]]))
    }
  }
  df
}

guess_clean <- prep_guess(test_guess_model_df, train_split)
train_no_y <- train_split |> select(-dep_del15)

# XGBoost guess matrix — built from training columns as reference
guess_xgb_mats <- make_aligned_matrix(
  ~ .,
  train_no_y,
  guess_clean
)
dguess <- xgb.DMatrix(data = guess_xgb_mats$x_test)
xgb_guess <- predict(xgb_model, dguess)

# Logistic regression — predicts directly from data frame, no matrix needed
log_reg_guess <- predict(log_reg, type = "response", newdata = test_guess_model_df)

# LASSO guess matrix
guess_lm_mats <- make_aligned_matrix(
  ~ . + poly(sched_dep_hour, 2),
  train_no_y,
  guess_clean
)
lasso_guess <- predict(cv_lasso, newx = guess_lm_mats$x_test,
                        s = "lambda.min", type = "response") |> as.vector()

# RF / Balanced RF
test_guess_rf <- guess_clean |>
  mutate(dep_del15 = factor(NA, levels = c("0", "1")))
test_guess_rf <- align_to_training(test_guess_rf, train_split_rf)
rf_guess <- predict(rf_model, newdata = test_guess_rf, type = "prob")[, "1"]
brf_guess <- predict(brf_model, newdata = test_guess_rf, type = "prob")[, "1"]

# Weighted ensemble
delay.guesses <- 0.50 * xgb_guess +
                 0.15 * rf_guess +
                 0.10 * brf_guess +
                 0.15 * lasso_guess +
                 0.10 * log_reg_guess

cat("Predictions:", length(delay.guesses),
    "| NAs:", sum(is.na(delay.guesses)), "\n")
```

```{r}
test_guess_fe |>
  summarise(
    pct_has_visible = mean(!is.na(visible_delay_rate)),
    mean_rate = mean(visible_delay_rate, na.rm = TRUE)
  )
```

```{r}
# Save submission file
team.name <- "Amador"
test.acc <- 1 - mean(vis_auc$accuracy[vis_auc$model_name == "Ensemble (Weighted)"])

save(list = c("delay.guesses", "test.acc", "team.name"),
     file = paste0(team.name, "462final.RData"))
```